---
title: "3: Physical Properties of Lakes"
author: "Hydrologic Data Analysis | Kateri Salk"
date: "Fall 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---
a
## Lesson Objectives
1. Investigate the concepts of lake stratification and mixing by analyzing monitoring data
2. Apply data analytics skills to applied questions about physical properties of lakes
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion
What is a lake? How does this differ from a stream, river, or wetland?
Lakes are bodies of water in a basin or depression. They are landlocked and by the ocean like lagoons. Flow is slow compared to rivers and streams. 
Many wetlands have emergent vegetation from sediment surface to water surface. Lakes are too deep for emergent vegetation. 
In the deepest part of a lake, you don't get light penetration all the way at the bottom. 

What are the physical properties of lakes?

## Session Set Up

getwd()
```{r, message = FALSE}
# Check working directory (should be project file location)

setwd("/Users/gabrielagarcia/Desktop/Hydrologic Data Analysis/Hydrologic_Data_Analysis/Data_Raw")
# load packages
library(tidyverse)
library(lubridate)


```

# Load data
```{r}
NTLdata <- read.csv("NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
```

# set ggplot theme
```{r}
theme_set(theme_classic())
```

## Data Wrangling and Exploration

### Investigate structure of dataset

Common steps/checks for data wrangling: 

* Date formatting
* Addressing NAs
* Subsetting and filtering 

# Is the date column perceived as a date by R?
```{r}
class(NTLdata$sampledate)
```

```{r}

NTLdata$sampledate <- as.Date(NTLdata$sampledate, "%m/%d/%y")
```

```{r}

class(NTLdata$sampledate)

```
###Do dates like the above format-year, month, day


# What does this column look like now?

###Check dimensions
```{r}
dim(NTLdata)
```


###Drop rows with no temperature measurement
```{r}
NTLdata <- NTLdata %>%
  drop_na(temperature_C)

```

###Check dimensions again
```{r}
dim(NTLdata)
```


# Remove rows that contain NAs in temperature column

## How many observations are there for each lake?
```{r}
summary(NTLdata$lakename)
```



# Let's choose the three lakes with data
```{r}
NTLdata <- NTLdata %>%
  filter(lakename %in% c("Paul Lake", "Peter Lake", "Tuesday Lake"))
```



# What is another way to use the filter command to get to the same result?
```{r}
NTLdata <- NTLdata %>%
  filter(lakename == "Paul Lake" | lakename == "Peter Lake" | 
           lakename == "Tuesday Lake")
```



# Make three data frames, one for each lake
```{r}
Pauldata <- filter(NTLdata, lakename == "Paul Lake")
Peterdata <- filter(NTLdata, lakename == "Peter Lake")
Tuesdaydata <- filter(NTLdata, lakename == "Tuesday Lake")
```



# How long did the monitoring last?
```{r}
min(Pauldata$sampledate)
max(Pauldata$sampledate)
min(Peterdata$sampledate)
max(Peterdata$sampledate)
min(Tuesdaydata$sampledate)
max(Tuesdaydata$sampledate)
```



# Which depths are sampled in each lake?
```{r}
unique(Pauldata$depth)
unique(Peterdata$depth)
unique(Tuesdaydata$depth)
```


# Why didn't we use the "summary" function here?
The summary function tells you the min, mean, max, and quartiles. The unique function lets us know the unique depths sampled. From R help, "unique returns a vector, data frame or array like x but with duplicate elements/rows removed."

### Exploratory data visualization

Let's make a plot of temperatures by depth. There are a lot of points, so adding a 50 % transparency to the points helps us see where points are densely clustered together.
```{r}
TempvsDepth <- 
  ggplot(Pauldata, aes(x = depth, y = temperature_C)) + 
  geom_point(alpha = 0.5) +
  labs(y = expression("Temperature "(degree*C)), x = "Depth (m)")
print(TempvsDepth)
```

How do temperatures at the surface compare to temperatures at the mid-depths and at the bottom?
<Temperatures at the surface are warmer than temperatures at the bottom of the lake. The range of temperatures decrease as depth increases, so less light is able to penetrate the lake with depth. The bottom of the lake isn't interacting with the atmosphere.>

Let's make a few data frames that include measurements from specific depths. We will choose dates only from the 2000s for ease of interpretation.

```{r}
Pauldata2000s <- filter(Pauldata, year4 > 1999)
Pauldata.surface <- filter(Pauldata2000s, depth == 0) 
Pauldata.2m <- filter(Pauldata2000s, depth == 2)
Pauldata.3m <- filter(Pauldata2000s, depth == 3)
Pauldata.4m <- filter(Pauldata2000s, depth == 4)
Pauldata.5m <- filter(Pauldata2000s, depth == 5)
Pauldata.6m <- filter(Pauldata2000s, depth == 6)
Pauldata.7m <- filter(Pauldata2000s, depth == 7)
Pauldata.9m <- filter(Pauldata2000s, depth == 9)
Pauldata.bottom <- filter(Pauldata2000s, depth == 12)
```

Now let's plot a few of the data frames on the same ggplot. How do temperatures at the surface compare to those at 2 m depth? How do temperatures at the bottom compare to those at 9 m depth?
<add notes here>

Here we are highlighting depths considered the **epilimnion** and **hypolimnion**.
```{r}
Tempplot.mixed <- 
  ggplot(Pauldata.surface, aes(x = sampledate, y = temperature_C)) +
  geom_point(color = "#6baed6", size = 1) + 
  geom_point(data = Pauldata.2m, aes(x = sampledate, y = temperature_C), 
             color = "#2171b5", size = 1) +
  geom_point(data = Pauldata.9m, aes(x = sampledate, y = temperature_C), 
             color = "#08519c", size = 1) +
  geom_point(data = Pauldata.bottom, aes(x = sampledate, y = temperature_C), 
             color = "#08306b", size = 1) +
  labs(y = expression("Temperature "(degree*C)), x = "Year")
print(Tempplot.mixed)
```
###The bottom layer of the lake is likely mixed-same temperatures-called the hypolimnion

###Surface layer of the lake is called the epilimnion
Now let's take out the 2 m and 9 m points and add in the 4 m and 6 m points. How do these compare to the surface and bottom temperatures?
<add notes here>

Here are are highlighting depths considered the **metalimnion** or **thermocline**.
```{r}
Tempplot.thermocline <- 
  ggplot(Pauldata.surface, aes(x = sampledate, y = temperature_C)) +
  geom_point(color = "#6baed6", size = 1) + 
  geom_point(data = Pauldata.4m, aes(x = sampledate, y = temperature_C), 
             color = "#2171b5", size = 1) +
  geom_point(data = Pauldata.6m, aes(x = sampledate, y = temperature_C), 
             color = "#08519c", size = 1) +
  geom_point(data = Pauldata.bottom, aes(x = sampledate, y = temperature_C), 
             color = "#08306b", size = 1) +
  labs(y = expression("Temperature "(degree*C)), x = "Year")
print(Tempplot.thermocline)
```
The graph above this one has separation between the surface and the bottom. This graph has no separation between surface and bottom. This


Why can't we use a line graph here?
<We're sampling from May to August-the data stops at August, and we start sampling again the next may, so a line graph wouldn't be the best for this time series data>

#Line graph
```{r}
Tempplot.line <- 
  ggplot(Pauldata.surface, aes(x = sampledate, y = temperature_C)) +
  geom_line(color = "#6baed6", size = 1) + 
  geom_line(data = Pauldata.4m, aes(x = sampledate, y = temperature_C), 
             color = "#2171b5", size = 1) +
  geom_line(data = Pauldata.6m, aes(x = sampledate, y = temperature_C), 
             color = "#08519c", size = 1) +
  geom_line(data = Pauldata.bottom, aes(x = sampledate, y = temperature_C), 
             color = "#08306b", size = 1) +
  labs(y = expression("Temperature "(degree*C)), x = "Year")
print(Tempplot.line)
```

## Data Visualization and Analysis
### Creating profile graphs

The field of **limnology**, the study of inland waters, uses a unique graph format to display relationships of variables by depth in a lake (the field of oceanography uses the same convention). Depth is placed on the y-axis in reverse order and the other variable(s) are placed on the x-axis. In this manner, the graph appears as if a cross section were taken from that point in the lake, with the surface at the top of the graph.

```{r}
Pauldata.2016 <- filter(Pauldata, year4 == 2016)
Pauldata.May2016 <- filter(Pauldata, sampledate == "2016-05-17")
Pauldata.June2016 <- filter(Pauldata, sampledate == "2016-06-21")
Pauldata.July2016 <- filter(Pauldata, sampledate == "2016-07-26")
Pauldata.Aug2016 <- filter(Pauldata, sampledate == "2016-08-16")
```


```{r}
TempprofileMay2016 <- 
  ggplot(Pauldata.May2016, aes(x = temperature_C, y = depth)) +
  geom_line(color = "#081d58") +
  geom_vline(xintercept = 4, lty = 2) + ###this line represents 4 degrees celsius-this is the temp of the maximum density of water. 4 degrees is the coldest a lake will get. 
  scale_y_reverse(breaks = c(0, 2, 4, 6, 8, 10, 12)) + ###this is to flip the scales
  scale_x_continuous(position = "top", limits = c(0, 30)) +
  labs(x = expression("Temperature "(degree*C)), y = "Depth (m)")
print(TempprofileMay2016)
```
###In May, the atmosphere and air warms, solar radiation heats up the lake at the surface. AS you go down in the water column however, the rest of the lake is cold and around 4 degrees celsius. Highest depth of the hypolimnion is about 5 meters. 

```{r}
TempprofileJune2016 <- 
  TempprofileMay2016 +
  geom_line(data = Pauldata.June2016, aes(x = temperature_C, y = depth), 
            color = "#253494")
print(TempprofileJune2016)
```


```{r}
TempprofileJuly2016 <- 
  TempprofileJune2016 +
  geom_line(data = Pauldata.July2016, aes(x = temperature_C, y = depth), 
            color = "#225ea8")
print(TempprofileJuly2016)
```


```{r}

TempprofileAug2016 <- 
  TempprofileJuly2016 +
  geom_line(data = Pauldata.Aug2016, aes(x = temperature_C, y = depth), 
            color = "#1d91c0")
print(TempprofileAug2016)
```



<add notes here>

In some places, the lines are not consistently smooth. What's going on here? How does this differ from your expectations of what the data should look like?
<The lines arent consistently smoothed because ecosystems aren't perfect. Measurements taken with a profiler, which might make slight errors below the water column>

If you were to explain this graph to someone who didn't know anything about lakes, how would you describe it? Write your answer below. 

> 

### Mixing and stratification

Let's visualize all of the sampled dates in 2016 at once.

```{r}

Tempprofiles2016 <- 
  ggplot(Pauldata.2016, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse() +
  scale_x_continuous(position = "top") +
  scale_color_viridis_c() +
  labs(x = expression("Temperature "(degree*C)), y = "Depth (m)", 
       color = "Julian Day")
print(Tempprofiles2016)
  
```

##Darkest colors are earliest dates, lightest colors are later dates. 
Paul Lake, like many lakes in temperate climates, is **dimictic**. What does this mean? What evidence do you see of this phenomenon in the graph? If you had data from the rest of the year, what would you expect to see?
<Dimictic means two and mixing, so this lake mixes twice throughout the year. Once during fall turnover; As air cools during fall, density gradient between top and bottom of lake decreases. This also happens during spring; Ice on top during winter will melt during spring, and you'll have a water column that's 4 degrees celsius all the way down.>


<Factors for Mixing:
-You need wind for diffusion
-You need a small enough density gradient so force of wind can mix up the water.>


Let's put your prediction to the test. In 1993, the lake was sampled in November. Wrangle your data to capture this date and create a profile graph of that date.
```{r}
November1993<-filter(Pauldata, year4 == 1993 | daynum == 306)

```

Why does mixing occur in the spring and fall? What are the mechanisms that make this possible?
<add notes here>

## Closing Discussion

What are the main concepts you learned about the physical properties of lakes today? What was the evidence for these concepts in the dataset?
