---
title: "4: Physical Properties of Lakes"
author: "Hydrologic Data Analysis | Kateri Salk"
date: "Fall 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## Lesson Objectives
1. Investigate the relationships between temperature, light, and oxygen in lakes by analyzing monitoring data
2. Analyze changes in lake surface temperatures over time
3. Apply data analytics skills to questions about physical properties of lakes
4. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion
We analyzed temperature profiles and seasonal changes on Wednesday. What were some key takeaway points about thermal stratification and mixing?


<epilimnion: well mixed surface layer. 
metalimnion: rapid change. Middle layer. 
hypolimnion: cold bottom layer that separated from the surface layer. Highest depth of the hypolimnion is about 5 meters.
We don't always have all of the layers present. Some scenarios for this would be a shallow, fully mixed lake. In addition, dimictic lakes spring and fall mixing because there is a smaller density gradient(and strong wind) that is able to overcome the density gradient to mix the lake.>
4 degrees celsius is the lower boundary of the hypolimnion. 
Liquid water is at its most dense at 4 degrees celsius. It is preparing to become ice. Once we hit 3 degrees and lower, the molecules begin to spread out and become less dense. If water decreases below 4 degrees celsius, it'll rise to the top.>

## Session Set Up
```{r, message = FALSE, warning=FALSE}
setwd("/Users/gabrielagarcia/Desktop/Hydrologic Data Analysis/Hydrologic_Data_Analysis/Data_Raw")

library(tidyverse)
library(gridExtra)
library(cowplot)
library(lubridate)

NTLdata <- read.csv("NTL-LTER_Lake_ChemistryPhysics_Raw.csv")

theme_set(theme_classic())

```

## Data Wrangling and Exploration

### Investigate structure of dataset
```{r}
# Change sampledate to Date
NTLdata$sampledate <- as.Date(NTLdata$sampledate, "%m/%d/%y") ###provide the format it came in rather than the date format we want
```

# Make three data frames, one for each lake
```{r}
Pauldata <- filter(NTLdata, lakename == "Paul Lake")
Peterdata <- filter(NTLdata, lakename == "Peter Lake")
Tuesdaydata <- filter(NTLdata, lakename == "Tuesday Lake")
```


# Make a data frame to inspect 2016 data for Paul Lake
```{r}
Pauldata.2016 <- filter(Pauldata, year4 == 2016)
```



### Data Visualization

Let's look at profiles of temperature, irradiance, and dissolved oxygen over the course of a season. We will use the `plot_grid` function from the cowplot package to put the graphs together on one plot. Pay attention to the differences among the formats of each plot and how these work together in the final plot. 

```{r}
Tempprofiles2016 <- 
  ggplot(Pauldata.2016, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse() +
  scale_x_continuous(position = "top") +
  scale_color_viridis_c(end = 0.8, option = "magma") + ###end=0.8 will not include yellow color
  labs(x = expression("Temperature "(degree*C)), y = "Depth (m)") +
  theme(legend.position = "none")
print(Tempprofiles2016)
```
###Faceting works if you have Peter Lake, Paul Lake, and Tuesday Lake (different layers within a column). 

```{r}
Lightprofiles2016 <- 
  ggplot(Pauldata.2016, aes(x = irradianceWater, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse() +
  scale_x_continuous(position = "top") +
  scale_color_viridis_c(end = 0.8, option = "magma") +
  labs(x = expression("Irradiance (W m"^-2*")"), y = "Depth (m)") +
  theme(legend.position = "none", 
        axis.text.y = element_blank(), axis.title.y = element_blank())
print(Lightprofiles2016)
```

```{r}
Oxygenprofiles2016 <- 
  ggplot(Pauldata.2016, aes(x = dissolvedOxygen, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse() +
  scale_x_continuous(position = "top") +
  scale_color_viridis_c(end = 0.8, option = "magma") +
  labs(x = expression("DO (mg L"^-1*")"), y = "Depth (m)", 
       color = "Julian Day") +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank())
print(Oxygenprofiles2016)
```


```{r, fig.width=4}
Physicalprofiles2016 <- 
  plot_grid(Tempprofiles2016, Lightprofiles2016, Oxygenprofiles2016, 
            ncol = 3, rel_widths = c(1.25, 1, 1.5))  #ncol=3 columns, relative widths can be specified
print(Physicalprofiles2016)

```
###Because all three graphs have the same y-axis data, we only need the y axis label on the left most y-axis
140=may, 160-june 9th, 180=june 28th, 200=july 18th, 220=august 7th



How do temperature profiles vary over the course of a season? What are the mechanisms for this pattern?

> Temperature profiles will be the lowest in May(140), and increase in June, July, and August throughout the course of a season due to high air temperatures and strong solar radiation. 

How do irradiance profiles vary over the course of a season? What are the mechanisms for this pattern?

> Irradiance will decrease after summer because solar radiation decreases, therefore the farther down in lake depth that is sampled, the less irradiance will be measured. Light levels decline below the surface because light is refracting off of the water molecules the further down the water column you go. The light penetration can also be impacted by algal blooms. They will absorb and scatter irradiance, so when you have a lot of biomass in water column, irradiance will decrease as you go down further in depth. Dissolved organic matter can also scatter light and impact irradiance (these lakes tend to be darker)

How do dissolved oxygen profiles vary over the course of a season? What are the mechanisms for this pattern?

> DO concentrations appears to decrease (albeit not a linear relationship) withh depth. You have higher dissolved oxygen concentrations in cold water (colder water has capacity to hold more dissolved gases), and lower dissolved oxygen concentrations in warmer water. There is a jump in DO at the metalimnion where dissolved oxygen concentrations decrease at about 5 meters in depth. In addiiton, photosynthesis increases oxygen concentration. Phytoplankton hang out at intersection of hypolimnion and metalimnion=called deep chlorophyll maximum (spike in chlorophyll maximums, or pigment used in photosynthesis). There is enough light to grow and enough nutrients. Another process that could affect oxygen concentrations are respiration. 




## Data Analysis

### Long-term change in temperature
How much have surface temperatures increased over the period of study? Is this increase significant? Isolate surface depths and run the test for May, June, July, and August temperatures separately. Use a linear regression with year as the predictor variable and temperature as the response variable.

Steps: 

1. Add a column named "Month" to the data frame (hint: lubridate package)
2. Filter your data frame so that it only contains surface depths and months 5-8
3. Create 4 separate data frames, one for each month
4. Run a linear regression for each data frame (see instructions above)
5. For significant trends, calculate how many degrees the lake has warmed over the period of study
5. Create a ggplot of surface temps by date and facet by month. Add `geom_smooth(se = FALSE, method = lm)` to the plot to display the regression line.
Tempprofiles2016

##Add Month Column to Pauldata data frame
```{r}

Pauldata$Month<-as.Date(Pauldata$sampledate,format="%m/%d/%y")



Pauldata<-mutate(Pauldata, Month=month(sampledate))

```

##Filter Pauldata frame so it contains only surface depths and months 5-8
```{r}
Pauldatafiltered<-filter(Pauldata, depth==0.0, Month==c(5, 6, 7, 8))
```

##PaulData for May
```{r}
PauldatafilteredMay<-filter(Pauldatafiltered, Month=="5")
```

##Paul Data for June
```{r}
PauldatafilteredJune<-filter(Pauldatafiltered, Month=="6")
```

##Paul Data for July
```{r}
PauldatafilteredJuly<-filter(Pauldatafiltered, Month=="7")
```

##Paul Data for August
```{r}
PauldatafilteredAugust<-filter(Pauldatafiltered, Month=="8")
```

##Linear Regression for May
```{r}
PaulLinearRegressionMay<-lm(PauldatafilteredMay$temperature_C~PauldatafilteredMay$year4)
summary(PaulLinearRegressionMay)

##
```


##Linear Regression for June
```{r}
PaulLinearRegressionJune<-lm(PauldatafilteredJune$temperature_C~PauldatafilteredJune$year4)
summary(PaulLinearRegressionJune)
```

##Linear Regression for July
```{r}
PaulLinearRegressionJuly<-lm(PauldatafilteredJuly$temperature_C~PauldatafilteredJuly$year4)
summary(PaulLinearRegressionJuly)

##Temperature increases 0.06 degrees per year(0.06*33) (33 is number of years data spans)
##198 degree increase over period of study for July
```

##Linear Regression for August
```{r}
PaulLinearRegressionAugust<-lm(PauldatafilteredAugust$temperature_C~PauldatafilteredAugust$year4)
summary(PaulLinearRegressionAugust)

##Temperature increases 0.04 degrees per year
0.04*33
#1.32 degree increase over period of study for August
```

###plot
```{r}
PaulPlot<- ggplot(Pauldatafiltered, aes(x = sampledate, y = temperature_C)) +
  geom_point()+
  geom_smooth(se=FALSE, method=lm)+
  facet_grid(rows=vars(Month))
print(PaulPlot)
```


Summarize your results below. How would you explain your findings and he importance of these findings to a non-specialist audience while still employing the data and statistical results?

>

## Closing Discussion
How do warming surface waters impact the physical properties of a lake? How might this affect the ecology of the lake?

How do we reconcile large amounts of variability in long-term records?
