---
title: "8: Water Quality in Rivers"
author: "Hydrologic Data Analysis | Cathy Chamberlin"
date: "Fall 2019"
output: html_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## Lesson Objectives
1. Analyze biological aspects of water quality.
2. Compare water quality data to hydrologic data.
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion
What are indicator species? Why are they helpful in thinking about water quality?
>If a species lives in a water body, it's likely tolerant for the existing water quality. 

## Session Set Up

```{r, message = FALSE}
setwd("/Users/gabrielagarcia/Desktop/Hydrologic Data Analysis/Hydrologic_Data_Analysis/Data_Raw")

packages <- c("tidyverse", 
              "dataRetrieval", 
              "cowplot", 
              "xts",
              "dygraphs", 
              "lubridate")
invisible(lapply(packages, library, character.only = TRUE)) 


theme_set(theme_classic(base_size = 8))
```

```{r}
gabytheme <- theme_bw(base_size = 14) + 
  theme(plot.title=element_text(face="bold", size="20", color="IndianRed4", hjust=0.5),
        axis.title=element_text(face="bold.italic", size=11, color="black"),
axis.text = element_text(face="bold", size=10, color = "black"), 
panel.background=element_rect(fill="gray96", color="darkblue"), 
panel.border = element_rect(color = "black", size = 2),
legend.position = "top", legend.background = element_rect(fill="white", color="black"),
            legend.key = element_rect(fill="transparent", color="NA"))
```

# Microbiological indicators

Bacteria is present in almost every stream on earth. Most bacteria are harmless, but some are extremely dangerous to human health.Typical sources of bacterial contamination include waste water treatment plants, animal manure lagoons, and septic systems. There are multiple ways that bacteria can be measured for water quality.

1) **Total cell counts**. One way to measure the bacteria in water is to count the number of cells under a microscope in a given volumneof water (such as 1 mL). The units for this metric are cells ml^-1^. You basically take a volume of water and pass cells through a narrow area and count them. 

2) **Total coliforms**. Coliform bacteria are for the most part harmless microorganisms that live in animal intestines. Total coliforms are often used as a metric of how much raw or insufficiently treated sewage is making its way into a water body. The way these are measured are by taking a known volume of water, cultivating it in the lab, and counting the number of colonies that grow. Units for this metric are colony forming units (cfu) per 100 ml.
3) **Fecal coliforms**. Fecal coliforms are coliforms that are specifically associated with warm-blooded animals, and can be grown at higher temperatures than other coliforms. Fecal coliforms are a more precise way of estimating sewage inputs, though they are only an indicator species and not necessarily harmful. The most well known species in this group is Escherichia coli (E. coli). Units are cfu per 100 ml. 
4) **Fecal streptocci**. Fecal streptocci are a different type of bacteria that are also associated with human and animal digestive tracts. Streptocci are more widespread than coliforms. Units are cfu per 100 ml.

Why might it be helpful to have multiple ways to measure bacteria?

>None of these methods are error-foolproof, so using multiple measuremnt methods will help you cross-validate the data. 

Where and when do you think the biggest risks from harmful bacteria are?

>

Measuring microbiogial indicators is more expensive and time consuming than some other water quality metrics. Where and when would you measure these?

>Low priority streams and rivers that aren't in the watershed could be sampled a lot less, but drinking water reservoirs would be sampled more frequently. Downstream wastewater treatment plants are also sampled frequently. 

# How have improvements in sewage treatment affected water quality?

We will import data from the Merrimack River near Lowell, Massachusetts. The Merrimack valley has a long history of industrial use, and has many mid-sized cities along the river corridor. Most of these cities have old infrastructure, and their wastewater treatment plants (WWTPs) are built next to the river and discharge directly into the water. Since the passage of the Clean Water Act, plants have either been retrofitted, or new facilities have been built to try to meet the criteria for the *point source discharges* (pipes that go out into streams) 

We will import a number of water quality parameters. Remember to reference ```parameterCdFile``` for metadata about the parameters.

```{r pull_data}
parameters <- parameterCdFile

dat.raw <- readNWISqw(siteNumbers = "01096550", #Merrimack River upstream of Lowell 
                     parameterCd = c("95200",  #Total cell count
                                     "31501",  #Total coliforms
                                     "31625",  #Fecal coliforms (0.7 um or 0.7 micron filter size, which will allow more bacteria to pass through, so the water will have more bacterial colonies)
                                   "31616",  #Fecal coliforms (0.45 um or 0.45 micron filter size)
                                     "31673",  #Fecal streptocci KF streptocaccus
                                     "31679"), #Fecal streptocci m-Enterococcus
                     startDate = "",
                     endDate = "")
```

### Look at Structure of Data
```{r}
str(dat.raw, give.attr = FALSE)
```

## Data Restructuring to Select Relevant Parameters 
```{r}
microb.dat <- dat.raw %>%
  select(Date = sample_dt, 
         Parameter = parm_cd, 
         Value = result_va) %>%
  group_by(Date, Parameter) %>%
  summarize(Value = mean(Value)) %>%
  spread(key = Parameter, value = Value) %>%
  rename(Total.coli = '31501',
         F.coli.45 = '31616', 
         F.coli.7 = '31625',
         F.strep.strep = '31673',
         F.strep.ente = '31679',
         Total.cell = '95200')

```

```{r}
begin.date <- min(microb.dat$Date)
end.date <- max(microb.dat$Date)
```

```{r}
Total.coli.plot <- ggplot(microb.dat, aes(x = Date)) +
  geom_line(aes(y = Total.coli)) +
  lims(x = c(begin.date, end.date)) +
  labs(x = "", y = "Total coliforms (cfu / 100 ml)") +
  theme(legend.position = "top")

print(Total.coli.plot)

```


```{r}
F.coli.45.plot <- ggplot(microb.dat, aes(x = Date)) +
  geom_line(aes(y = F.coli.45)) +
  lims(x = c(begin.date, end.date)) +
  labs(x = "", 
       y = expression("Fecal coliforms filtered to 0.45 " * mu *m *  " (cfu / 100 ml)")) 

print(F.coli.45.plot)
```



```{r}

F.coli.7.plot <- ggplot(microb.dat, aes(x = Date)) +
  geom_line(aes(y = F.coli.7)) +
  lims(x = c(begin.date, end.date)) +
  labs(x = "", 
       y = expression("Fecal coliforms filtered to 0.7 " * mu *m *  " (cfu / 100 ml)")) 

print(F.coli.7.plot)
```


```{r}

F.strep.strep.plot <- ggplot(microb.dat, aes(x = Date)) +
  geom_line(aes(y = F.strep.strep)) +
  lims(x = c(begin.date, end.date)) +
  labs(x = "", 
       y = "Fecal streptocci, KF streptococcus (cfu / 100 ml)") 

print(F.strep.strep.plot)
```


```{r}
F.strep.ente.plot <- ggplot(microb.dat, aes(x = Date)) +
  geom_line(aes(y = F.strep.ente)) +
  lims(x = c(begin.date, end.date)) +
  labs(x = "", 
       y = "Fecal streptocci, m-Enterococcus (cfu / 100 ml)") 

print(F.strep.ente.plot)
```


```{r}

Total.cell.plot <- ggplot(microb.dat, aes(x = Date)) +
  geom_line(aes(y = Total.cell), show.legend = FALSE) +
  lims(x = c(begin.date, end.date)) +
  labs(x = "", y = expression("Total cell count (# ml"^-1*")")) 

print(Total.cell.plot)
```


```{r =fig.width=15, fig.height=20}
Microbes.plot <- plot_grid(Total.coli.plot +
                             labs(title = "Total Coliform", y = ""),
                           F.coli.45.plot + 
                             labs(title = "Fecal Coliform (.45)", y = "") ,
                           F.coli.7.plot + 
                             labs(title = "Fecal Coliform (.7)", y = "") ,
                           F.strep.strep.plot + 
                             labs(title = "Fecal streptocci (KF streptococcus)",
                                  y = ""),
                           F.strep.ente.plot +
                             labs(title = "Fecal streptocci (m-Enterococcus)",
                                  y = ""), 
                           Total.cell.plot + 
                             labs(title = "Total cell count", y = ""),
                           ncol = 1)+
  theme(plot.title=element_text(size=10))+
  gabytheme

print(Microbes.plot)
```


How would you use all of this data to talk about changes in microbial water quality from 1970 to 2000? What are the challenges?

>The scales on the graphs are all different so its hard to compare the different bacterias. We also don't have data for the entire period of record. 
The Total Coliforms includes the fecal coliform 0.7 and  0.45 (fecal coliforms are some percentage of total coliforms)
If we measured all three types, we'd have the highest amount of bacterial cooinies in Total Coliforms, less colonies with 0.7, and the lowest amount of colonies with 0.45 filter. 

>The change in microbial bacteria has decreased. The fecal coliform (0.45) amounts from 1970 -1980 would result in less colonies than the fecal coliform (0.7) amounts from 1970-1980, if there was actually data for the fecal coliform (0.7). Even if we can't say exactly how much the amount of colonies have decreased, we know the general trend is decreasing. 

>

Do you think water quality has improved between 1970 and 2000? Why or why not?

>

The city of Lawrence, MA is downstream of Lowell and gets its drinking water from the Merrimack River. Use the federal drinking water standard for Total Coliforms (https://www.epa.gov/ground-water-and-drinking-water/national-primary-drinking-water-regulations#Microorganisms) to comment on whether or not this water necessarily needs to be treated for bacteria.



# Seasonal Trends in Water Quality

## Add Date Column 
```{r seasons_analysis}
monthly.obs <- microb.dat %>%
  mutate(month = month(Date)) %>%
  group_by(month) %>%
  select(-Date) %>%
  summarize_all(funs(sum(!is.na(.))))

```

##Calculate average 
```{r}
coli.monthly.summaries <- microb.dat %>%
  mutate(month = month(Date)) %>%
  group_by(month) %>%
  select(month, Total.coli) %>%
  summarize_all(funs(Median = median(., na.rm = T),
                 quant25 = quantile(., .25, na.rm = T),
                 quant75 = quantile(., .75, na.rm = T))) 
```

```{r}
coli.seasons <- ggplot(coli.monthly.summaries, aes(x = month)) +
  geom_ribbon(aes(ymin = quant25, ymax = quant75), alpha = 0.3) +
  geom_line(aes(y = Median)) +
  scale_x_continuous(name = "Month",
                     breaks = c(1,2,3,4,5,6,7,8,9,10,11,12)) +
    labs(y = "Total coliforms (cfu / 100 ml)") 

print(coli.seasons)
```




When are total coliform concentrations highest? Why might this be?

>Colfiromns are highest in July, August, and September and you have minimums in the winter months. This is because warmer summer temperatures allow microbes to grow more prolifically. It's not a perfect correlation (we see some time lag in November and December). 

Does this pattern also exist for the other microbial indicators?

## F Coli 45
```{r}
fcoli.plot <- ggplot(microb.dat, aes(x = Date)) +
  geom_line(aes(y = F.coli.45, show.legend = FALSE))+
  lims(x = c(begin.date, end.date)) +
  labs(x = "", y = expression("Total cell count (# ml"^-1*")"))

print(fcoli.plot)

```

```{r}
Fcoli.plot7 <- ggplot(microb.dat, aes(x = Date)) +
  geom_line(aes(y = F.coli.7, show.legend = FALSE))+
  lims(x = c(begin.date, end.date)) +
  labs(x = "", y = expression("Total cell count (# ml"^-1*")"))

print(Fcoli.plot7)
```

```{r}

```


## Combined Sewer Overflows (CSOs)

Many older WWTPs combine the sanitary sewer system of a city and the storm water drainage sewers (e.g. street drains) together before treatment. Treatment occurs in tanks dug in the facilities, and treatment takes usually about a day. When there is a big rainfall event, the amount of water coming through the storm drains can overwhelm the holding capacity in the tanks, and untreated sewage can get discharged into receiving waterbodies. We will next look to see if any of the high bacterial measurements correspond to storm events.

The USGS site that we got the water quality data from does not have discharge data. Other nearby sites with discharge data include a site (```01100000```) 4 miles downstream on the Merrimack River, or a site (```01099500```) 4 miles away on the Concord River. The Merrimack River at this location drains ~4,000 square miles, and the Concord River (about 4 miles away from Merrimack river) drains ~400 square miles. The Concord River gage will be a better indicator since it'll react much more to lcoal storm conditions since it's a smaller area and drains a smaller watershed. 

>In wastewater teatment plants, water drains out of storm water system (street)and from residences (municipal water system). If it rains a lot, the amount of water coming out of the storm drain also increases, so water can't spend all day sitting in a fixed size tank and musr be pushed out. As stormwater is leaving tank, it takes municipal water with it. 
What would microbial indicators look like if raw sewage entered the stream? The amount of microbial indicators would increase. The discharge after the storm event would be high

Which site's discharge data would better indicate when the city of Lowell has experienced a big rain event?

>

What would a CSO event look like in the data?

>

## Pull in Hydrology Daily Data
```{r pull_hydrology}
discharge.raw <- readNWISdv(site = "01099500", parameterCd = c("00060"), startDate = begin.date, endDate = end.date)
str(discharge.raw, give.attr = FALSE)
```

## Select columns wanted 
```{r}
discharge.dat <- discharge.raw %>%
  select(Date = Date, Discharge_cfs = X_00060_00003)
```

```{r}
  
ggplot(discharge.dat, aes(x = Date, y = Discharge_cfs)) +
  geom_line() +
  labs(y = "Discharge (cfs)", x = "Date")
```


We will use an interactive plotting function, dygraphs to look at the discharge data and microbial data overlain on each other. Let's look first at Total Coliform.

```{r make_dygraph)}
Total_Coliform <- with(microb.dat, xts(x = Total.coli, order.by = Date))  
Discharge <- with(discharge.dat, xts(x = Discharge_cfs, order.by = Date))###with discharge data, we create xts where x=discharge, and we're going to order by the Date

DyDat <- cbind(Total_Coliform, Discharge) ### bind the columns together to create a matrix
```

```{r, fig.width=8}
dygraph(DyDat) %>% 
  dySeries("Total_Coliform", axis = "y2") %>% 
  dyAxis(name = "y", label = "Discharge (cfs)") %>%
  dyAxis(name = "y2", label = "Total Coliform (cfu per 100 ml)", valueRange = c(0, 600000)) %>%
  dyRangeSelector()
```
>If we just ran dygraph(DyDat) without the other code below it, it would just give us discharge data and microbial data. The scales are extremely off though, so we put Total Colfiorm on a different y-axis. Thus, discharge and coliforms are on two different y-axes. Use dyaxis to give the axes different names and labels
We can also specify the value range, which allows us to  


>look for high value of microbial counts and some increase in discharge or change =when a CSO occurred


Look through the data series and see if you can find some times when a CSO likely occured. Create a ggplot of the 3-4 weeks surrounding the event. (Hint: use the skeleton code available below )
>August 13, 1976
>August 7, 1972 (decreasing discharge with super high fecal coliform values), 


High coliform with high discharge=coliform washed
High coliform during falling discharge occurs during point source pollution coming into the system. When natural discharge from watershed decreases, there is a higher proportion of point source discharge in the water.
Also, if flow isn't going very fast, the amount of coliform grows.

```{r}

```







```{r CSO_event, eval=FALSE}

begin.date <- 
end.date <- 
  
discharge.plot <- ggplot(data =  %>% 
                   filter(), 
                 aes(x = , y = )) +
            geom_line() +
            scale_x_date(limits =  , name  = )
  
microb.plot <- ggplot(data =  %>% 
                   filter(), 
                 aes(x = )) +
            geom_point(aes(y = ))+
            scale_x_date(limits =  , name = )
  
  
plot_grid( , , ncol = 1)

```

Show your plot to your group and discuss why you think it might represent a CSO.

>

## Closing Discussion

We have looked at 2 different measures of water quality this week. What are some of the other important parameters we could measure for water quality? Where would we be most concerned about them?

>

# References

Merrimack River - A Comprehensive Watershed Assessment Report (Massachusetts Executive Office of Environmental Affairs): https://www.mass.gov/files/2017-07/Merrimack%20River%20Watershed%20Assessment%20Report.pdf

National Primary Drinking Water Regulations (EPA): https://www.epa.gov/ground-water-and-drinking-water/national-primary-drinking-water-regulations

