---
title: "11: Time Series Analysis"
author: "Hydrologic Data Analysis | Kateri Salk"
date: "Fall 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## Lesson Objectives
1. Discuss the purpose and application of time series analysis for hydrologic data
2. Decompose time series into individual components

## Session Set Up

```{r, message = FALSE}
getwd()
setwd("/Users/gabrielagarcia/Desktop/Hydrologic Data Analysis/Hydrologic_Data_Analysis/Lessons")

library(tidyverse)
library(dataRetrieval)

theme_set(gabytheme)
```

## Time Series Analysis

Time series are a special class of dataset, where a response variable is tracked over time. The frequency of measurement and the timespan of the dataset can vary widely. At its most simple, a time series model includes an explanatory time component and a response variable. Mixed models can include additional explanatory variables (check out the `nlme` and `lme4` R packages). We will be covering a few simple applications of time series analysis in these lessons.

### Opportunities

Analysis of time series presents several opportunities. In aquatic sciences, some of the most common questions we can answer with time series modeling are:

* Has there been an increasing or decreasing **trend** in the response variable over time?
* Can we **forecast** conditions in the future?


### Challenges

Time series datasets come with several caveats, which need to be addressed in order to effectively model the system. A few common challenges that arise (and can occur together within a single dataset) are: 

* **Autocorrelation**: Data points are not independent from one another (i.e., the measurement at a given time point is dependent on previous time point(s))
>Example: If we tried to predict air temeprature tomorrow if we knew today's air temp: we know there's a correlation between today's temp and tomorrow's temp: autocorrelation. Our sample points are not independent if we sample them too close together. We often need to account for autocorrelation within a time series model.

* **Data gaps**: Data are not collected at regular intervals, necessitating *interpolation* between measurements.
There are often gaps between monitoring periods. For many time series analyses, we need equally spaced points. 

* **Seasonality**: Cyclic patterns in variables occur at regular intervals, impeding clear interpretation of a monotonic (unidirectional) trend. Ex. We can assume that summer temperatures 

* **Heteroscedasticity**: The variance of the time series is not constant over time


* **Covariance**: the covariance of the time series is not constant over time.
Many of these models assume that the variance and covariance are similar over the time-->heteroschedasticity. 


## Visualizing a time series dataset

Today, we will analyze discharge data from Clear Creek in Colorado, USA. Let's first look at what types of data are available for this dataset. 
```{r}
ClearCreekSummary <- whatNWISdata(siteNumbers = "06719505")
```
Notice that mean daily discharge has been measured at this site since 1974, with over 16,000 measurements available. This will be a robust time series dataset for us to analyze changes that have occurred over the past 45 years. 

Let's gather all the discharge data available from this site. Notice that `readNWISdv` gathers dates, which are already formatted as a date object. What do you notice about the data when plotted over time?

```{r}
# Import data
ClearCreekDischarge <- readNWISdv(siteNumbers = "06719505",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "",
                     endDate = "")

```


```{r}
names(ClearCreekDischarge)[4:5] <- c("Discharge", "Approval.Code")
class(ClearCreekDischarge$Date)
```

# Learn about the variables and site

```{r}
attr(ClearCreekDischarge, "variableInfo")
attr(ClearCreekDischarge, "siteInfo")
```


# Plot discharge over time
```{r, fig.width=3}
ClearCreekPlot <- 
  ggplot(ClearCreekDischarge, aes(x = Date, y = Discharge)) +
  geom_line() +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)")) + 
  theme(plot.title = element_text(margin = margin(b = -10), size = 12), 
       axis.title.x = element_blank())
print(ClearCreekPlot)

```
>This data appears to be cyclical, as we see peaks occurring at regular intervals. Cyclical data is typically on larger timescales. El Nino and La Nina Cycles show up between every 2-8 years. There doesn't seem to be a trend over time though. 
Watch out for the clear variance by season. 

The simplest option to characterize a time series is a linear regression. 
```{r, fig.width=6}
ClearCreekRegressionPlot <- 
  ggplot(ClearCreekDischarge, aes(x = Date, y = Discharge)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE, color = "#c13d75ff") +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)")) + 
  theme(plot.title = element_text(margin = margin(b = -10), size = 12), 
       axis.title.x = element_blank())
print(ClearCreekRegressionPlot)
```

Is a linear regression an appropriate model for the data in this case? Why or why not?

> It is not an appropriate model for the data in this case. 

Are there situations where it would be appropriate to use a linear regression to model a time series? If so, what is an example?

>If we had a seasonal cycle and we wanted to look at August 12th of each year, then we'd have an eqaul frequency of sampling events that we could apply a linear regression to. If we don't have seasonality to our data, that would be better to use in a linear regression.
We often decompose data into annnual means-->we see the change in means over the years-->this is when we can use a linear regression. 
-->Try to choose the simplest model that makes the most sense for our data. 


## Decomposing a time series dataset

A given time series can be made up of several component series: 

1. A **seasonal** component, which repeats over a fixed known period (e.g., seasons of the year, months, days of the week, hour of the day). 

2. A **trend** component, which quantifies the upward or downward progression over time. The trend component of a time series does not have to be monotonic.
This doesn't need to be unidirectional: it could go up for a little bit, and go down again.

3. An **error** or **random** component, which makes up the remainder of the time series after other components have been accounted for. This component reflects the noise in the dataset. 
Anything not explained by the seasons will be classified as error-we'll alwys have randomness in our data incorporated into our model. 

4. (optional) A **cyclical** component, which repeats over periods greater than the seasonal component. A good example of this in hydrologic data is El Niño Southern Oscillation (ENSO) cycles, which occur over a period of 2-8 years.

We first need to turn the discharge data into a time series object in R. This is done using the `ts` function. Notice we can only specify one column of data and need to specify the period at which the data are sampled. The resulting time series object cannot be viewed like a regular data frame.
>So if we want to provide discharge, what info are we not including? Sample date. 
>We can't include dates 

Note: time series objects must be equispaced, which requires interpolation if the data have not been collected at regular intervals. In our case, we have daily data with no NAs in the data frame, so we don't need to worry about this.


### Check the dates to see if there's a gap in sampling dates
```{r}
table(diff(ClearCreekDischarge$Date))
```
>Apparantly there is a gap, but we're going to ignore it for the sake of the lesson 

```{r}
ClearCreek_ts <- ts(ClearCreekDischarge[[4]], frequency = 365)

```
>Because there's no information stored, we need to make sure the timestep is the same. If there were NA's in our dataset, we'd need to get rid of them!
>We say frequency=365 because we're l
>Notice how this is stored as a value and not a dataframe


The `stl` function decomposes the time series object into its component parts. We must specify that the window for seasonal extraction is either "periodic" or a specific number of at least 7. The decomposition proceeds through a loess (locally estimated scatterplot smoothing) function.

```{r}
?stl
```
>Loess takes 

# Generate the decomposition
```{r}
ClearCreek_Decomposed <- stl(ClearCreek_ts, s.window = "periodic")
```
>This is just a list

# Visualize the decomposed series. 
```{r}
plot(ClearCreek_Decomposed)
```
>Notice the trend: It's notincreasing constantly. This is a moving trend, or the moving average across the dataset. 
>Remainder-->The error not predicted by seasonal and trend. If we added seasonal + trend + remainder, this equals our data.
>The highest discharge event occurs after 20 years

# We can extract the components and turn them into data frames
```{r}
ClearCreek_Components <- as.data.frame(ClearCreek_Decomposed$time.series[,1:3]) ###Pull in the timeseries list to create a dataframe
ClearCreek_Components <- mutate(ClearCreek_Components,
                      Observed = ClearCreekDischarge$Discharge,     
                      Date = ClearCreekDischarge$Date) ##add in date so we can look at these over time
```



# Visualize how the trend maps onto the data
```{r}
ggplot(ClearCreek_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Discharge (ft"^3*"/s)"))
```
>In years with higher discharge, our trend tends to be higher (ex. around the mid 90s, around 2010). For this dataset, the trend is always positive. 


# Visualize how the seasonal cycle maps onto the data
```{r}

ggplot(ClearCreek_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Discharge (ft"^3*"/s)"))
```
>There are negative seasonal discharge values. This is ok, but we need to pay attention to how it's impacting our data. We might also end up with numbers that are not realistic-our final predicted numbers must make sense. If we add up trend+seasonal+error, this results in the actual observed data. 

Note that the decomposition can yield negative values when we apply a seasonal adjustment or a trend adjustment to the data. The decomposition is not constrained by a lower bound of zero as discharge is in real life. Make sure to interpret with caution!

## Closing Discussion
Next lesson we will discuss how to backcast, forecast, and detect monotonic trends in time series data.
