---
title: "Assignment 5: Water Quality in Lakes"
author: Gaby Garcia
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Hydrologic Data Analysis on water quality in lakes

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single HTML file.
8. After Knitting, submit the completed exercise (HTML file) to the dropbox in Sakai. Add your last name into the file name (e.g., "A05_Salk.html") prior to submission.

The completed exercise is due on 2 October 2019 at 9:00 am.

# Setup

1. Verify your working directory is set to the R project file, 
2. Load the tidyverse, lubridate, and LAGOSNE packages. 
3. Set your ggplot theme (can be theme_classic or something else)
4. Load the LAGOSdata database and the trophic state index csv file we created on 2019/09/27.

### Set Working Directory
```{r}
setwd("/Users/gabrielagarcia/Desktop/Hydrologic Data Analysis/Hydrologic_Data_Analysis/Assignments")
```

### Load Packages
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
install.packages("LAGOSNE")
library(LAGOSNE)
library(corrplot)
library(cowplot)
library(dplyr)
```


###Set GGPlot Theme
```{r}
gabytheme <- theme_bw(base_size = 14) + 
  theme(plot.title=element_text(face="bold", size="20", color="IndianRed3", hjust=0.5),
        axis.title=element_text(size=13, color="black"),
axis.text = element_text(face="bold", size=10, color = "black"), 
panel.background=element_rect(fill="white", color="darkblue"), 
panel.border = element_rect(color = "black", size = 2),
legend.position = "top", legend.background = element_rect(fill="white", color="black"),
            legend.key = element_rect(fill="transparent", color="NA"))

theme_set(gabytheme)
```

### Remove Scientific Notation from Data Frame
```{r}
options(scipen = 100) ###To remove scientific notation from data frame columns
```


### Load LAGOSdata database
```{r}
###LAGOSdata <- lagosne_load()
###names(LAGOSdata)

# If the package installation and data download has not worked, use this code: 
LagosneLoad<-load(file = "/Users/gabrielagarcia/Desktop/Hydrologic Data Analysis/Hydrologic_Data_Analysis/Data/Raw/LAGOSdata.rda")
```

##Obtain Metadata, State, and Nutrient Information
```{r}
LAGOSlocus <- LAGOSdata$locus
LAGOSstate <- LAGOSdata$state
LAGOSnutrient <- LAGOSdata$epi_nutr
```


# Tell R to treat lakeid as a factor, not a numeric value
```{r}
LAGOSlocus$lagoslakeid <- as.factor(LAGOSlocus$lagoslakeid)
LAGOSnutrient$lagoslakeid <- as.factor(LAGOSnutrient$lagoslakeid)
```

# Join data frames using a left join
```{r}
LAGOSlocations <- left_join(LAGOSlocus, LAGOSstate, by = "state_zoneid")
```

# Trophic State Index

5. Similar to the trophic.class column we created in class (determined from TSI.chl values), create two additional columns in the data frame that determine trophic class from TSI.secchi and TSI.tp (call these trophic.class.secchi and trophic.class.tp). 

## Create LAGOStrophic data frame from LAGOSnutrient and LAGOSlocations dataframes
```{r, warning=FALSE, message=FALSE}
LAGOStrophicfinal<- 
  left_join(LAGOSnutrient, LAGOSlocations, by = "lagoslakeid") %>%  ###left join of lagosnutrient and LAGOS locations by lagoslakeid
  
  select(lagoslakeid, sampledate, chla, tp, tn, secchi, 
         gnis_name, lake_area_ha, state, state_name) %>% ###select relevant columns 
  
  mutate(sampleyear = year(sampledate), 
         samplemonth = month(sampledate), 
         
         season = as.factor(quarter(sampledate, fiscal_start = 12))) %>% ###Creating a season column. Quarter function divides your year into 4 quarters of three months. Here we start it at month 12, so winter is dec, january, feb, spring is march, april, may, and so on. We want these numbers to be considered factors. 
  
  
  drop_na(chla:secchi) ###removing NA's from columns chla: secchi. If there is a NA in the row, it'll remove the entire row. 
```

## Add Trophic Indices columns for Secchi and Phosphorus
```{r}
LAGOStrophicfinal <- 
  mutate(LAGOStrophicfinal, 
         TSI.chl = round(10*(6 - (2.04 - 0.68*log(chla)/log(2)))),
         TSI.secchi = round(10*(6 - (log(secchi)/log(2)))), 
         TSI.tp = round(10*(6 - (log(48/tp)/log(2)))), 
         trophic.class=ifelse(TSI.chl < 40, "Oligotrophic", 
                   ifelse(TSI.chl < 50, "Mesotrophic",
                          ifelse(TSI.chl < 70, "Eutrophic", "Hypereutrophic"))), 
         trophic.class.secchi=ifelse(TSI.secchi < 40, "Oligotrophic", 
                   ifelse(TSI.secchi < 50, "Mesotrophic",
                          ifelse(TSI.secchi < 70, "Eutrophic", "Hypereutrophic"))),
         trophic.class.tp=ifelse(TSI.tp < 40, "Oligotrophic", 
                   ifelse(TSI.tp < 50, "Mesotrophic",
                          ifelse(TSI.tp < 70, "Eutrophic", "Hypereutrophic"))))
```




6. How many observations fall into the four trophic state categories for the three metrics (trophic.class, trophic.class.secchi, trophic.class.tp)? Hint: `count` function.



## Chorophyll Observations
```{r}
with(LAGOStrophicfinal, table(trophic.class))
```

## Secchi Observations
```{r}
with(LAGOStrophicfinal, table(trophic.class.secchi))
```

## Total Phosphorus Observations
```{r}
with(LAGOStrophicfinal, table(trophic.class.tp))
```

7. What proportion of total observations are considered eutrohic or hypereutrophic according to the three different metrics (trophic.class, trophic.class.secchi, trophic.class.tp)?


```{r}

```

Which of these metrics is most conservative in its designation of eutrophic conditions? Why might this be?

> 

Note: To take this further, a researcher might determine which trophic classes are susceptible to being differently categorized by the different metrics and whether certain metrics are prone to categorizing trophic class as more or less eutrophic. This would entail more complex code.

## Nutrient Concentrations


8. Create a data frame that includes the columns lagoslakeid, sampledate, tn, tp, state, and state_name. Mutate this data frame to include sampleyear and samplemonth columns as well. Call this data frame LAGOSNandP. 

```{r}
LAGOSNandP<-LAGOStrophicfinal%>%select(lagoslakeid, sampledate, tn, tp, state, state_name)%>% mutate(sampleyear = year(sampledate), 
         samplemonth = month(sampledate))
```



9. Create two violin plots comparing TN and TP concentrations across states. Include a 50th percentile line inside the violins.

# TN Concentrations
```{r, warning=FALSE, message=FALSE, fig.width=4}
TNViolin<-ggplot(LAGOSNandP, 
                           aes(x =factor(state), y =tn)) +
  geom_violin(draw_quantiles = c(0.5),
              aes(fill=factor(state))) +
  labs(title="Total Nitrogen Concentrations across the United States", 
       y="Discharge (ft^3/s)")
print(TNViolin)
```

# TP Concentrations
```{r}

```



Which states have the highest and lowest median concentrations?

> TN:

> TP:

Which states have the highest and lowest concentration ranges?

> TN: 

> TP: 

10. Create two jitter plots comparing TN and TP concentrations across states, with samplemonth as the color. Choose a color palette other than the ggplot default.

```{r}

```

Which states have the most samples? How might this have impacted total ranges from #9?

> TN:

> TP: 

Which months are sampled most extensively? Does this differ among states?

> TN:

> TP: 

11. Create two jitter plots comparing TN and TP concentrations across states, with sampleyear as the color. Choose a color palette other than the ggplot default.

```{r}

```


Which years are sampled most extensively? Does this differ among states?

> TN:

> TP: 

## Reflection
12. What are 2-3 conclusions or summary points about lake water quality you learned through your analysis?

>

13. What data, visualizations, and/or models supported your conclusions from 12?

> 

14. Did hands-on data analysis impact your learning about water quality relative to a theory-based lesson? If so, how?

>

15.	How did the real-world data compare with your expectations from theory?

>
